A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-07-22 paper/main.tex
saving revision progress

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls, we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a huge benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g., Participants P1, P4, and P15), while others' recalls segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations in participant's recalls.  Whereas each event in the original video (was largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory participants appear to be integrating \textit{across} different events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event~\citep[Figs.~\ref{fig:model}D, \corrmats; also see][]{MannEtal11, HowaEtal12}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original video, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for 2 of the analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature. Namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with a forward assymetric bias (Fig.~\ref{fig:list-learning}C). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed somewhat evenly throughout the video.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed 2 novel metrics for quantifying naturalistic memory representations: precision and distinctiveness.  We define precision as the average correlation between each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) as well as the number of recall events estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similar to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$ and model derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting Information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall shape of what happened), whereas the more minor details are prone to pruning.

In addition to constructing overall summaries, assessing the video and recall topic vectors from individual recalls can provide further insights.  Specifically, for any given event we can construct two wordles: one from the original video event's topic vector, and a second from the average topic vectors produced by all participants' recalls of that event.  We can then examine those wordles visually to gain an intuition for which aspects of the video event were recapitulated in participants' recalls of that event.  Several example wordles are displayed in Figure~\ref{fig:topics}C (wordles from the three best-remembered events are circled in green; wordles from the three worst-remembered events are circled in red).  Using wordles to visually compare the topical content of each video event and the (average) corresponding recall event reveals the specific content from the specific events that is reliably retained in the transformation into memory (green events) or not (red events).

@@

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportion matrix (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal cortex and cingulate cortex, suggesting that these regions may play a role in maintaining information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  We used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\ref{fig:model}D, \corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex (vmPFC), anterior cingulate cortex, and right medial temporal lobe (rMTL), suggesting that these regions may play a role in transforming each individual's experience into memory.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory analyses}~\citep[e.g.,][]{PallWagn02} to domain of naturalistic stimuli.

>>>>

Given that the time-varying content of the video could be segmented cleanly into discrete events, we wondered whether participants' recalls of the video also displayed a similar structure.  We applied the same topic model (already trained on the video annotations) to each participant's recalls.  Analogous to how we analyzed the time-varying content of the video, to obtain similar estimates for participants' recalls, we treated each (overlapping) 10 sentence ``window'' of their transcript as a ``document'' and then computed the most probable mix of topics reflected in each timepoint's sentences.  This yielded, for each participant, a number-of-sentences by number-of-topics topic proportions matrix that characterized how the topics identified in the original video were reflected in the participant's recalls.  Note that an important feature of our approach is that it allows us to compare participant's recalls to events from the original video, despite that different participants may have used different language to describe the same event, and that those descriptions may not match the original annotations.  This is a substantial benefit of projecting the video and recalls into a shared ``topic'' space.  An example topic proportions matrix from one participant's recalls is shown in Figure~\ref{fig:model}D.

@@

Two clear patterns emerged from this set of analyses.  First, although every individual participant's recalls could be segmented into discrete events (i.e., every individual participant's recall correlation matrix exhibited clear block diagonal structure; Fig.~\corrmats), each participant appeared to have a unique \textit{recall resolution}, reflected in the sizes of those blocks.  For example, some participants' recall topic proportions segmented into just a few events (e.g., Participants P1, P4, and P15), while others' recalls segmented into many shorter duration events (e.g., Participants P12, P13, and P17).  This suggests that different participants may be recalling the video with different levels of detail-- e.g., some might touch on just the major plot points, whereas others might attempt to recall every minor scene or action.  The second clear pattern present in every individual participant's recall correlation matrix is that, unlike in the video correlation matrix, there are substantial off-diagonal correlations in participant's recalls.  Whereas each event in the original video was (largely) separable from the others (Fig.~\ref{fig:model}B), in transforming those separable events into memory participants appear to be integrating \textit{across} different events, blending elements of previously recalled and not-yet-recalled events into each newly recalled event~\citep[Figs.~\ref{fig:model}D, \corrmats; also see][]{MannEtal11, HowaEtal12}.

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original video, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for 2 of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}C). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed somewhat evenly throughout the video.

% One may reasonably hypothesize that the diminished effect of presentation order on recall likelihood stems from 1) differences between the presence of certain features of episodic experiences that are absent in the simple, discretized stimuli of classic list-learning experiments.  Random word lists most often employ simple, common words While they are well-suited to the random word-list paradigm, these analyses address only the temporal order of stimuli, neglecting to consider their content (or, for example, that content's potential interrelatedness).  This difference serves to highlight the need for alternative approaches to quantifying memory success that account for such contextual features and the way they may impact memory organization and retrieval dynamics.

@@

Statistical models of memory studies often treat memory recalls as binary (e.g. the item was recalled or not) and independent events.  However, our framework produces a content-based model of individual stimulus and recall events, allowing for direct quantitative comparison between all stimulus and recall events, as well as between the recall events themselves.  Leveraging these content-based models of the stimulus/recall events, we developed two novel metrics for quantifying naturalistic memory representations:~\textit{precision} and \textit{distinctiveness}.  We define precision as the average correlation between the topic proportions of each recall event and the maximally correlated video event (Fig.~\ref{fig:precision-distinctiveness}).  Participants whose recall events are more veridical descriptions of what happened in the video event will presumably have higher precision scores. We find that across participants, a higher precision score is correlated to both hand annotated memory performance (Pearson's $r(15) = 0.6, p = 0.011$) and the number of recall events estimated by our model (Pearson's $r(15) = 0.64, p = 0.005$). A second novel metric we introduce here is distinctiveness, or how unique the recall description was to each video event. We define distinctiveness as 1 minus the average of all non-matching recall events from the video-recall correlation matrix. We hypothesized that participants who recounted events in a more distinctive way would display better overall memory.  Similarly to precision, we find that the more distinct participants recalls are (on average), the more they remembered (hand-annotated memory: Pearson's $r(15) = 0.83, p < 0.001$; model-derived memory: Pearson's $r(15) = 0.71, p = 0.001$).  In summary, using two novel metrics afforded by our approach, we find that participants whose recalls are both more precise and distinct remember more content.

@@

Because our analysis framework projects the dynamic video content and participants' recalls onto a shared topic space, and because the dimensions of that space are known (i.e., each topic dimension is a set of weights over words in the vocabulary; Fig.~\topics), we can examine the topic trajectories to understand which specific content was remembered well (or poorly).  For each video event, we can ask: what was the average correlation (across participants) between the video event's topic vector and the closest matching recall event topic vectors from each participant? This yields a single correlation coefficient for each video event, describing how closely participants' recalls of the event tended to reliably capture its content (Fig.~\ref{fig:topics}A).  (We also examined how different comparisons between each video event's topic vector and the corresponding recall event topic vectors related to hand-annotated characterizations of memory performance; see \textit{Supporting Information}).  Given this summary of which events were recalled reliably (or not), we next asked whether the better-remembered or worse-remembered events tended to reflect particular topics.  We computed a weighted average of the topic vectors for each video event, where the weights reflected how reliably each event was recalled.  To visualize the result, we created a ``wordle'' image~\citep{MuelEtal18} where words weighted more heavily by better-remembered topics appear in a larger font (Fig.~\ref{fig:topics}B, green box).  Events that reflected topics weighting heavily on characters like ``Sherlock'' and ``John'' (i.e., the main characters) and locations like ``221b Baker Street'' (i.e., a major recurring location and the address of the flat that Sherlock and John share) were best remembered.  An analogous analysis revealed which themes were poorly remembered.  Here in computing the weighted average over events' topic vectors, we weighted each event in \textit{inverse} proportion to how well it was remembered (Fig.~\ref{fig:topics}B, red box).  This revealed that events with relatively minor characters such as ``Mike,'' ``Jeffrey,'' and ``Molly,'' as well as less-integral plot locations (e.g., ``hospital'' and ``office'') were least well-remembered.  This suggests that what is retained in memory are the major plot elements (i.e., the overall shape of what happened), whereas the more minor details are prone to pruning.

In addition to constructing overall summaries, assessing the video and recall topic vectors from individual events can provide further insights.  Specifically, for any given event we can construct two wordles: one from the original video event's topic vector, and a second from the average topic vectors produced by all participants' recalls of that event.  We can then examine those wordles visually to gain an intuition for which aspects of the video event were recapitulated in participants' recalls.  Several example wordles are displayed in Figure~\ref{fig:topics}C (wordles from the three best-remembered events are circled in green; wordles from the three worst-remembered events are circled in red).  Using wordles to visually compare the topical content of each video event and the (average) corresponding recall event reveals the specific content from the specific events that is reliably retained in the transformation into memory (green events) or not (red events).

@@

The results thus far inform us about which aspects of the dynamic content in the episode participants watched were preserved or altered in participants' memories of the episode.  We next carried out a series of analyses aimed at understanding which brain structures might implement these processes.  In one analysis we sought to identify which brain structures were sensitive to the video's dynamic content, as characterized by its topic trajectory.  Specifically, we used a searchlight procedure to identify the extent to which each cluster of voxels exhibited a timecourse of activity (as the participants watched the video) whose temporal correlation matrix matched the temporal correlation matrix of the original video's topic proportions (Fig.~\ref{fig:model}B).  As shown in Figure~\ref{fig:brainz}A, the analysis revealed a network of regions including bilateral frontal cortex and cingulate cortex, suggesting that these regions may play a role in processing information relevant to the narrative structure of the video.  In a second analysis, we sought to identify which brain structures' responses (while viewing the video) reflected how each participant would later \textit{recall} the video.  We used an analogous searchlight procedure to identify clusters of voxels whose temporal correlation matrices reflected the temporal correlation matrix of the topic proportions for each individual's recalls (Figs.~\ref{fig:model}D, \corrmats).  As shown in Figure~\ref{fig:brainz}B, the analysis revealed a network of regions including the ventromedial prefrontal cortex (vmPFC), anterior cingulate cortex (ACC), and right medial temporal lobe (rMTL), suggesting that these regions may play a role in transforming each individual's experience into memory.  In identifying regions whose responses to ongoing experiences reflect how those experiences will be remembered later, this latter analysis extends classic \textit{subsequent memory analyses}~\citep[e.g.,][]{PallWagn02} to domain of naturalistic stimuli.
